---
title: 'Genome coverage: Jens vs CSD datasets'
author: "Cyril Matthey-Doret"
date: "July 11, 2018"
output: pdf_document
---

After observing a very small overlap between Jens linkage map markers and the wasps from the CSD project, I ran his samples through my pipeline to check if only SNPs differ or if we sequence different regions of the genome (e.g. due to differences in protocols/conditions). I compare the relative number of loci overlapping between the dataset Jens used to anchor the genome (via a linkage map) and the CSD dataset. A small overlap would be indicative of a difference in protocols whereas a large overlap would be normal.

Conclusions: Similar genomic regions are sequenced, there is no real difference in genome coverage, which indicate there is no issue with differences in protocols. Jens data has much fewer SNPs, which makes sense, as he only has a single non introgressed family while we have several families post-introgression. Together, these infos mean we cannot make a linkage map using Jens data and we would need to make one from the same introgressed families we have (which is not possible). Next step: Extrapolate cM from BP using Jens data and apply the transformation to our dataset. This is pretty bad, but the best we can do.

```{bash interval, echo=F}
DATADIR="comp_jens_CSD"
# Generate bed files from CSTACKS output
for f in csd jens;do
  zcat "$DATADIR/$f.catalog.tags.tsv.gz" \
    | tail -n +2 \
    | awk -v fname="$f" 'BEGIN{OFS="\t"}{print $4,$5,$5+94,0,fname"_"$3,$6}' \
    | sort -k1,1 -k2,2n > "$DATADIR/$f.bed"
done

# compute intersections
cd "$DATADIR"
bedtools intersect -wo -a "csd.bed" -b "jens.bed" > "intersect.bed"
```

```{r intersect_stats, echo=F}
library(readr);library(dplyr);library(eulerr);library(gridExtra)
# Compare intersection vs total coverages
isect <- read_tsv('comp_jens_CSD/intersect.bed', 
                  col_names = F, col_types = cols())

jens <- read_tsv('comp_jens_CSD/jens.bed', 
                 col_names = F, col_types = cols()) 
# %>% filter(! X5 %in% isect$X11)

csd <- read_tsv('comp_jens_CSD/csd.bed', 
                col_names = F, col_types = cols()) 
# %>% filter(! X5 %in% isect$X5)


jens_snp <- read_tsv('comp_jens_CSD/jens.sumstats.tsv', 
                     comment = '#', col_names = F, 
                     col_types = cols()) %>%
  mutate(ID=paste(X3, X4, sep='_'))

csd_snp <- read_tsv('comp_jens_CSD/csd.sumstats.tsv', 
                    comment = '#', col_names = F, 
                    col_types = cols()) %>%
  mutate(ID=paste(X3, X4, sep='_'))

isect_snp <- csd_snp %>% inner_join(jens_snp, by = 'ID')

#csd_snp <- csd_snp %>% filter(! ID %in% isect_snp$ID)
#ens_snp <- jens_snp %>% filter(! ID %in% isect_snp$ID)
```

```{r plot_results}
# Visualise intersection sets
euler_loci <- plot(euler(c('jens'=nrow(jens),'csd'=nrow(csd),'jens&csd'=nrow(isect)), shape = 'circle'), quantities = T)
grid.arrange(grobs = list(euler_loci), top = "Number of overlapping loci")

# Show units as Mb
zoom <-10^6
euler_bp <- plot(euler(c('jens'=nrow(jens)*94/zoom,
                         'csd'=nrow(csd)*94/zoom,
                         'jens&csd'=sum(isect$X13)/zoom), shape = 'circle'), quantities = T)
grid.arrange(grobs = list(euler_bp), top = "Size of overlap [Mb]")

euler_snp <- plot(euler(c('jens'=nrow(jens_snp),
                         'csd'=nrow(csd_snp),
                         'jens&csd'=nrow(isect_snp)), shape = 'circle'), quantities = T)
grid.arrange(grobs = list(euler_snp), top = "Number of overlapping SNPs")

```